---
title: "Homework 2"
author: "Tara Zhan"
data: 09-26-2023
output: github_document
---

```{r, message=FALSE, echo=FALSE}
library(moderndive)
library(tidyverse)
library(readxl)
library(haven)
library(readxl)
library(dplyr)
```

# Problem 1

```{r, echo=FALSE}
# Read three CSV files:
pols <- read.csv("fivethirtyeight_datasets/pols-month.csv")
snp <- read.csv("fivethirtyeight_datasets/snp.csv")
unemployment <- read.csv("fivethirtyeight_datasets/unemployment.csv")
```

```{r}
# First, clean the data in 'pols-month.csv'.
pols_cleaned <- pols %>%
  # Separate the `mon` column into `year`, `month`, and `day`
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>%
  # Replace month number with month name
  mutate(month = month.abb[as.numeric(month)]) %>%
  # Create a `president` variable
  mutate(president = ifelse(!is.na(prez_dem), "dem", ifelse(!is.na(prez_gop), "gop", NA))) %>%
  # Remove `prez_dem`, `prez_gop`, `day` variables
  select(-day, -prez_dem, -prez_gop)
```

**pols_cleaned/(pols-month.csv)**: This dataset describes the voting rate of presidents from 1947 to 2015. The cleaned version contains variables like 'year', 'month', 'political party of the president' that either Democratic ('dem') or republican ('gop').

```{r}
# Second, clean the data in 'snp.csv'
snp_cleaned <- snp %>%
  # Separate the `date` column into `year`, `month`, and `day'
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>%
  # Replace month number with month name
  mutate(month = month.abb[as.numeric(month)]) %>%
  # Arrange the data by year and month
  arrange(year, month) %>%
  # Organize 'year' and 'month' are the leading columns
  select(year, month, everything(), -day)
```

**snp_cleaned/(snp.csv)**: This dataset describes the stock market indexes over time. The cleaned version arranged data chronologically by 'year' and 'month', making the dataset easier to read. It also contains 'close' values.

```{r}
# Third, tidy the 'unemployment.csv' then merged with the previous datasets
unemployment_long <- unemployment %>%
  #convert from wide to long format
  gather(key = "month", value = "unemployment_rate", -Year)
unemployment_long <- unemployment_long %>%
  rename(year = Year)
```

**unemployment_cleaned/(unemployment.csv)**: This dataset describes unemployment rates in each months from 1948 to 2015. The tidying version convert the dataset from wide to long format, making it easier to merge with other datasets. It also contains 'unemployment rate' values.

```{r}
# Merging 'snp' into 'pols', and merging 'unemployment' into the result.
combined_data <- pols_cleaned %>%
  # Merge 'snp' into 'pols'
  left_join(snp_cleaned, by = c("year", "month")) %>%
  # Merge 'unemployment'
  left_join(unemployment_long, by = c("year", "month"))
```

**combined_data:** 
This is a combination of three above cleaned version datasets based on the 'year' and 'month'. This combined dataset describes an outlook of political, economic, and unemployment rates from year 1948 to year 2015. This combined_data has 11 columns, 822 rows, and spans from the year 1947 to 2015. Key variables in this dataset include: year, month, gov_gop, sen_gop, rep_gop, gov_dem, sen_dem, rep_dem, president, close, unemployment_rate.  
```{r, echo=FALSE, eval=FALSE}
print(str(combined_data))
ncol(combined_data)
nrow(combined_data)
min(combined_data$year)
max(combined_data$year)
paste(names(combined_data), collapse=", ")
```

# Problem 2
```{r}
# clean Mr. Trash Wheel sheet, and add a new variables
mr_trash_wheel <- 
  read_excel("trash_wheel_collection.xlsx", 1, range="A2:N549") |>
  janitor::clean_names() |>
  filter(homes_powered!=0.00000) |>
# based on calculation.
  mutate(homes_powered = weight_tons * (500/30))
view(mr_trash_wheel)
```

```{r}
#clean Professor Trash Wheel sheet 
prof_trash_wheel <- 
  read_excel("trash_wheel_collection.xlsx", 2, range="A2:M96") |>
  janitor::clean_names() |>
  drop_na() |>
# based on calculation.
  mutate(homes_powered = weight_tons * (500/30))
view(prof_trash_wheel)
```

```{r}
#clean Gwynnda Trash Wheel sheet 
gwyn_trash_wheel <- 
  read_excel("trash_wheel_collection.xlsx", 4, range="A2:K108") |>
  janitor::clean_names() |>
  drop_na() |>
# based on calculation.
  mutate(homes_powered = weight_tons * (500/30))
view(gwyn_trash_wheel)
```

```{r}
# Add wheel_name
mr_trash_wheel <- mr_trash_wheel %>%
  mutate(wheel_name = "Mr. Trash Wheel")
prof_trash_wheel <- prof_trash_wheel %>%
  mutate(wheel_name = "Professor Trash Wheel")
gwyn_trash_wheel <- gwyn_trash_wheel %>%
  mutate(wheel_name = "Gwynnda Trash Wheel")
# Convert year to numeric
mr_trash_wheel$year <- as.numeric(mr_trash_wheel$year)
prof_trash_wheel$year <- as.numeric(prof_trash_wheel$year)
gwyn_trash_wheel$year <- as.numeric(gwyn_trash_wheel$year)

#combine all datasets
combined_trash_wheel <- bind_rows(mr_trash_wheel, prof_trash_wheel, gwyn_trash_wheel)
# View the combined dataset
view(combined_trash_wheel)
```

The "combined_trash_wheel" dataset is a combination of data derived from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel. This combination includes a total of `r nrow(combined_trash_wheel)` observations, and calculate the number of homes powered. It is computed using the formula provided, where each ton of trash is equated to 500 kilowatts of electricity.  
For available data, the total weight of trash collected by Professor Trash Wheel is `r sum(prof_trash_wheel$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(gwyn_trash_wheel$cigarette_butts[gwyn_trash_wheel$year == 2021 & gwyn_trash_wheel$month == "July"])` cigarette butts.

# Problem 3

```{r, message=FALSE}
# Read CSV files:
baseline <- 
  read.csv("data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  rename("study_id" = "id")
# Convert sex and APOE4 status to factor 
  baseline %>%
    mutate(
      sex = recode(sex, "1" = "male", "0" = "female"),
      apoe4 = recode(apoe4, "1" = "carrier", "0" = "non-carrier")
      )
# delete no MCI at baseline
baseline$age_at_onset <- 
  replace(baseline$age_at_onset, baseline$age_at_onset == ".", NA)
cleaned_baseline <- 
  print(baseline) |>
  drop_na() 
cleaned_baseline$age_at_onset <- as.numeric(cleaned_baseline$age_at_onset)
```
After importing and reformatting the dataset, the "sex" and "APOE4" variables were converted to meaningful characters, and participants who did not develop MCI were removed from the dataset.
At the beginning, this study recruited `r nrow(baseline)` participants, and among them only `r nrow(cleaned_baseline)` participants develop MCI. Based on the data from cleaned version, the average baseline age is around `r mean(cleaned_baseline$age_at_onset)` years, and based on the total participants, the proportion of women are carriers is around `r sum(baseline$sex == "female" & baseline$apoe4 == "carrier")/sum(baseline$sex == "female") * 100`% people. 

```{r}
# Read CSV files:
amyloid <- 
  read.csv("data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |>
  drop_na()
# convert to long format
amyloid_long <- amyloid %>%
  pivot_longer(baseline:time_8, 
               names_to = "time", 
               values_to = "value")
```
After importing and reformatting the dataset, all "NA" values were deleted, and the dataset was convert into long format.
There are `r length(setdiff(baseline$id, amyloid_long$study_id))` participants appear in only the baseline dataset, and `r length(setdiff(amyloid_long$study_id, baseline$id))` participants appear in only the amyloid dataset.
```{r}
# Merge datasets
combined_mci <- merge(baseline, amyloid_long, by = "study_id" )
# Exporting the combined_mci to a CSV file
write.csv(combined_data, "data_mci/combined_dataset.csv", row.names = FALSE)
```
The "combined_mci" dataset is a combination of data derived from baseline and amyloid_long. This combination includes both demographic details and observed biomarker values, easier to analysis Alzheimer’s disease.
